---
title: ""
format: html
---

# 🐧 Palmer Penguins demo using `dplyr`

The **`dplyr`** package provides a set of simple, consistent functions for manipulating data frames.

We’ll explore some of the most common ones for choosing columns, rows and grouping variables to perform summary statistics.

---

## Load data

```{r}
#| message: false

library(dplyr)
library(palmerpenguins)
head(penguins)
```

```{r}
# Optional: force penguins to appear in your environment:
penguins <- penguins
```


## what are tibbles?

stuff goes here!

## Select columns

Columns use `select()` function

```{r}
penguins |>
  select(species, island, body_mass_g)
```

## Filter rows

Rows use `filter()` function


```{r}
penguins |>
  filter(body_mass_g > 4000)
```

## create new columns 

```{r}
penguins |>
  mutate(body_mass_kg = body_mass_g / 1000)
```

## summarise and group by

tend to use together

`group_by()` doesn’t change the data immediately — it just tells R how to organise it.

first this is what group by does

```{r}
penguins_grouped <- penguins |>
  group_by(species)
penguins_grouped
```

then this is summarise:

```{r}
penguins_grouped |>
  summarise(
    mean_flipper = mean(flipper_length_mm, na.rm = TRUE),
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    n = n()
  )
```

In practice, you’ll often combine them in one pipeline:

```{r}
penguins |>
  group_by(species) |>
  summarise(
    mean_flipper = mean(flipper_length_mm, na.rm = TRUE),
    mean_mass = mean(body_mass_g, na.rm = TRUE)
  )
```

## Putting it all together

Above we did each manipulation step by step. But eh beauty of dplyr and using teh pipe is puttign it altogether in one line of code.

Allows us to avoid making a new object for each step.

```{r}
# Select only relevant columns
# Filter to include only penguins heavier than 4000 g
# Create a new column converting grams to kilograms
# Group data by species
# Summarise the grouped data

penguins |>
  select(species, island, flipper_length_mm, body_mass_g) |> 
  filter(body_mass_g > 4000) |>   
  mutate(body_mass_kg = body_mass_g / 1000) |> 
  group_by(species) |>   
  summarise(
    mean_flipper = mean(flipper_length_mm, na.rm = TRUE),
    mean_mass_kg = mean(body_mass_kg, na.rm = TRUE),
    n = n()
  )   
```

Lastly, lets visualise it!

```{r}
library(ggplot2)

penguins_summary <- penguins |>
  select(species, island, flipper_length_mm, body_mass_g) |> 
  filter(body_mass_g > 4000) |>   
  mutate(body_mass_kg = body_mass_g / 1000) |> 
  group_by(species) |>   
  summarise(
    mean_flipper = mean(flipper_length_mm, na.rm = TRUE),
    mean_mass_kg = mean(body_mass_kg, na.rm = TRUE),
    n = n()
  )

ggplot(penguins_summary, aes(x = species, y = mean_mass_kg, fill = species)) +
  geom_col() +
  labs(title = "Mean body mass of penguins by species",
       y = "Mean body mass (kg)", x = "Species") +
  theme_minimal()
```


### Add error bars

Our bar plot looks ok, but we don't have any error bars. Lets do some more data manipulation to get standard deviation and standard error. 

Note, we need to use our original penguins dataset here, as we are goign to calucate the.... 

```{r}
penguins_summary_stats <- penguins |>
  select(species, island, flipper_length_mm, body_mass_g) |> 
  filter(body_mass_g > 4000) |>   
  mutate(body_mass_kg = body_mass_g / 1000) |> 
  group_by(species) |>   
  summarise(
    mean_mass_kg = mean(body_mass_kg, na.rm = TRUE),
    sd_mass_kg   = sd(body_mass_kg, na.rm = TRUE),   # standard deviation
    n            = n(),
    se_mass_kg   = sd_mass_kg / sqrt(n)             # standard error
  )
```

Now replot:

```{r}
ggplot(penguins_summary_stats, aes(x = species, y = mean_mass_kg, fill = species)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_mass_kg - se_mass_kg,
                    ymax = mean_mass_kg + se_mass_kg),
                width = 0.2) +  # width of error bar caps
  labs(title = "Mean Body Mass of Penguins with SE",
       y = "Mean Body Mass (kg)", x = "") +
  theme_minimal() +
  theme(legend.position = "none")
```

::: {.callout-note collapse="true" appearance="simple"}
# Plot with SD instead

```{r}
ggplot(penguins_summary_stats, aes(x = species, y = mean_mass_kg, fill = species)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_mass_kg - sd_mass_kg,
                    ymax = mean_mass_kg + sd_mass_kg),
                width = 0.2) +  # width of error bar caps
  labs(title = "Mean Body Mass of Penguins with SD",
       y = "Mean Body Mass (kg)", x = "") +
  theme_minimal() +
  theme(legend.position = "none")
```

:::

# Summary of key dplyr functions

- `select()` = choose **columns** (variables)
- `filter()` = choose **rows** (observations) that meet a condition
- `mutate()` = create or modify **columns**
- `group_by()` = define **groups** for summarising
- `summarise()` = calculate **summary statistics** per group